School Socioeconomic Status Context and Social Skills in Children

### 2020-10-28 manuscript.Rmd
Minor edits to manuscript

---
title             : "School Socioeconomic Status Context and Social Skills in Children"
shorttitle        : "School SES and Social Skills"
author: 
  - name          : "Philip Parker"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "33 Berry St, North Sydney, NSW, 2060, Australia"
    email         : "philip.parker@acu.edu.au"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Taren Sanders"
    affiliation   : "1"
    corresponding : no    # Define only one corresponding author
    address       : "33 Berry St, North Sydney, NSW, 2060, Australia"
    email         : "taren.sanders@acu.edu.au"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "Institute for Positive Psychology and Education, Australian Catholic University"
  - id            : "2"
    institution   : "UCL Institute of Education, UCL"
authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
  Enter author note here.
abstract: |

  As researchers and policymakers try to understand the consequences of SES for educational and occupational attainment, they increasingly consider social skills to be critical. Yet little research has considered how the school socio-economic context may differentially promote or hinder social skill development. In a representative longitudinal sample of 5361 Australian children, we tested the association between school average socioeconomic status and social skills as reported by parents and teachers. All models controlled for prior social skills at age 4, and a wide range of covariates. We found that school context was associated with social skill assimilation: controlling for individual socioeconomic status, children in more advantaged schools had more prosocial behavior and fewer peer and conduct problems. We found a consistent interaction between individual and school average socioeconomic status that suggested assimilation effects were only present for children from low socioeconomic backgrounds. We found that children from low socioeconomic backgrounds enter school with lower average social skills. They are more likely to be enrolled in more disadvantaged schools. Finally, the social skills of children from low socioeconomic backgrounds worsen in low SES contexts. Taking the evidence together, social stratified schools may harm children from low SES backgrounds while not benefiting children from high SES backgrounds.
keywords          : "social skills; assimilation effects; socioeconomic status; school context"
wordcount         : "X"
bibliography      : ["r-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : 
  pdf: 
    papaja::apa6_pdf
  html_document: 
    fig_caption: yes
    toc: yes
  html_notebook: 
    fig_caption: yes
    toc: yes
    highlight: haddock
    theme: spacelab
  word:
    papaja::apa6_doc
---

@@

We use data for children, their parents, and their teachers from the B-Cohort and K-Cohort of the Longitudinal Study of Australian Children (LSAC). LSAC is a government-run study of a representative sample of Australian children between who were zero-one (B-cohort) and four-five (K-Cohort) years of age in 2004. Both cohorts of children have been followed every two years since [@aifs2015]. LSAC includes linked administration records for student performance in a standardized national numeracy test. We also use government collected data on school average socioeconomic status. Many children in the sample were the only child surveyed (`r school_one`) or were one of two children surveyed (`r school_two`) in their school. By using administrative records, we gain school average socioeconomic status based on the child's $\text{\underline{whole}}$ school grade from a high-quality source, thus avoiding sampling bias present in much school context research. We excluded participants who were not in school (e.g., were home schooled, or were otherwise missing an identifier of which school they attended), were not in Year 3 at the time of testing, or were not eligible to complete national standardized testing (e.g., children with an intellectual disability). Together our total sample was `r nrow(data)` (`r boys` boys) children in Year 3 at school (\~age 8). In the vast majority of cases (98%), parent data came from mothers. A participant flow diagram is present end in Figure \@ref(fig:consort) including the number of participants excluded because there were a) not in Year 3 during the age 8 LSAC data collection sweep; b) were not eligible to participate NAPLAN (e.g., due to intellectual disability; or c) were not enrolled in school (e.g., were home schooled).

@@

All models in the paper that predicted social skills were estimated taking into account the censoring of the social skills means using a Bayes version of Tobit regression. Tobit models can be used when seeking to model outcome variables that are censored on the left, right, or both tails ([see @kleiber2008]. As scores on the social skills variables have both a floor of zero and a ceiling of 10, we included censoring on both the right and left in all models. The multilevel nature of the data (children nested within schools) was addressed by the inclusion of a random intercept for schools. We clearly demarcate the units of analysis in tables and figures. I general the SDQ measures are retained on the original scale and all other continuous variables are z-scored. We retain the SDQ scores on the original scale because of the presence of norms [\@youth2016scoring]. Peer and conduct problems scores of 3 or greater and prosocial scores of 7 or lower place children beyond the 'close to average' group and may thus be a potential concern [@youth2016scoring]. <!--# We could use post-stratification weights to provide clinically relevant proportions of the population for different levels of SES and school SES -->

@@

$$
y^{*}_{i} \sim N(\alpha_{j[i]} + X_i\beta + \sigma^{2}_{y}), for~i = 1,...,n\\
\alpha_{j} \sim N(U_j \gamma, \sigma^{2}_{\alpha}), for~j = 1,...k
$$

@@

$$
y_i =  \begin{cases}      y*_{i} & \text{if $y*_{i} < 0$ or $y*_{i} > 10 $}\\      0 & \text{if $y*_{i} < 0$}\\      10 & \text{if $y*_{i} > 10$}    \end{cases}
$$

@@

Parker et al. [-@parker2016; -@parker2018 ; -@parker_information_2018; -@parker2019] have suggested that empirical evidence shows greater school choice at the country level is related to poorer average ability levels, lower aspirations, and paradoxical effects on psychological factors like motivation and self-concept that appears to have negative consequences for all children. They argue that empirical evidence suggests policy should encourage school selection policies that maximize within school heterogeneity. This would require considerable state intervention to achieve and may thus impose unreasonable restrictions on parents' rights to choose. However, it is worth noting that our counterfactual analysis (Table 4) shows that high SES children appear to not suffer to any notable degree, in relation to social skills, by being enrolled in a disadvantaged school.

>>>>

---
title: "School Socioeconomic Status Context and Social Skills in Children"
author:

- name: Philip Parker
  affiliation: '1'
  corresponding: yes
  address: 33 Berry St, North Sydney, NSW, 2060, Australia
  email: philip.parker@acu.edu.au
  role:
  - Conceptualization
  - Formal Analysis
  - Writing - Original Draft Preparation
  - Writing - Review & Editing
- name: Taren Sanders
  affiliation: '1'
  # corresponding: no
  # address: 33 Berry St, North Sydney, NSW, 2060, Australia
  # email: taren.sanders@acu.edu.au
  role:
  - Conceptualization
  - Writing - Original Draft Preparation
  - Writing - Review & Editing
- name: Michael Noetel
  affiliation: '1,3'
  # corresponding: no
  # address: 33 Berry St, North Sydney, NSW, 2060, Australia
  # email: taren.sanders@acu.edu.au
  role:
  - Conceptualization
  - Writing - Original Draft Preparation
  - Writing - Review & Editing
shorttitle: School SES and Social Skills
output:
  pdf: papaja::apa6_pdf
  html_document:
    fig_caption: yes
    toc: yes
  pdf_document:
    toc: yes
  word: papaja::apa6_doc
  html_notebook:
    fig_caption: yes
    toc: yes
    highlight: haddock
    theme: spacelab
authornote: |
abstract: |2
  As researchers and policymakers try to understand the consequences of SES for educational and occupational attainment, they increasingly consider social skills to be critical. Yet little research has considered how the school socio-economic context may differentially promote or hinder social skill development. In a representative longitudinal sample of 5361 Australian children, we tested the association between school average socioeconomic status and social skills as reported by parents and teachers. All models controlled for prior social skills at age 4, and a wide range of covariates. We found that school context was associated with social skill assimilation: controlling for individual socioeconomic status, children in more advantaged schools had more prosocial behavior and fewer peer and conduct problems. We found a consistent interaction between individual and school average socioeconomic status that suggested assimilation effects were only present for children from low socioeconomic backgrounds. We found that children from low socioeconomic backgrounds enter school with lower average social skills. They are more likely to be enrolled in more disadvantaged schools. Finally, the social skills of children from low socioeconomic backgrounds worsen in low SES contexts. Taking the evidence together, social stratified schools may harm children from low SES backgrounds while not benefiting children from high SES backgrounds.
keywords: social skills; assimilation effects; socioeconomic status; school context
wordcount: X
bibliography: r-references.bib
floatsintext: no
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
documentclass: apa6
classoption: man
affiliation:
- id: '1'
  institution: Institute for Positive Psychology and Education, Australian Catholic
    University
- id: '2'
  institution: UCL Institute of Education, UCL
- id: '3'
  institution: School of Health and Behavioural Sciences, Australian Catholic University
---

@@

We use data for children, their parents, and their teachers from the B-Cohort and K-Cohort of the Longitudinal Study of Australian Children (LSAC). LSAC is a government-run study of a representative sample of Australian children between who were zero-one (B-cohort) and four-five (K-Cohort) years of age in 2004. Both cohorts of children have been followed every two years since [@aifs2015]. LSAC includes linked administration records for student performance in a standardized national numeracy test. We also use government collected data on school average socioeconomic status. Many children in the sample were the only child surveyed (`r school_one`) or were one of two children surveyed (`r school_two`) in their school. By using administrative records, we gain school average socioeconomic status based on the child's $\text{\underline{whole}}$ school grade from a high-quality source, thus avoiding sampling bias present in much school context research. We excluded participants who were not in school (e.g., were home schooled, or were otherwise missing an identifier of which school they attended), were not in Year 3 at the time of testing, or were not eligible to complete national standardized testing (e.g., children with an intellectual disability). Together our total sample was `r nrow(data)` (`r boys` boys) children in Year 3 at school (\~age 8). In the vast majority of cases (98%), parent data came from mothers. A participant flow diagram is presented in Figure \@ref(fig:consort) including the number of participants excluded because there were a) not in Year 3 during the age 8 LSAC data collection sweep; b) were not eligible to participate NAPLAN (e.g., due to intellectual disability; or c) were not enrolled in school (e.g., were home schooled).

@@

All models in the paper that predicted social skills were estimated taking into account the censoring of the social skills means using a Bayes version of Tobit regression. Tobit models can be used when seeking to model outcome variables that are censored on the left, right, or both tails [see @kleiber2008]. As scores on the social skills variables have both a floor of zero and a ceiling of 10, we included censoring on both the right and left in all models. The multilevel nature of the data (children nested within schools) was addressed by the inclusion of a random intercept for schools. We clearly demarcate the units of analysis in tables and figures. I general the SDQ measures are retained on the original scale and all other continuous variables are z-scored. We retain the SDQ scores on the original scale because of the presence of norms [@youth2016scoring]. Peer and conduct problems scores of 3 or greater and prosocial scores of 7 or lower place children beyond the 'close to average' group and may thus be a potential concern [@youth2016scoring]. <!--# We could use post-stratification weights to provide clinically relevant proportions of the population for different levels of SES and school SES -->

@@

$$
\begin{split}
y^{*}_{i} \sim N(\alpha_{j[i]} + X_i\beta + \sigma^{2}_{y}), for~i = 1,...,n \\
\alpha_{j} \sim N(U_j \gamma, \sigma^{2}_{\alpha}), for~j = 1,...k
\end{split}
$$

@@

$$
\begin{split}
y_i =  \begin{cases}      
            y*_{i} & \text{if $y*_{i} < 0$ or $y*_{i} > 10 $}\\      
            0 & \text{if $y*_{i} < 0$}\\      
            10 & \text{if $y*_{i} > 10$}    
        \end{cases}
\end{split}
$$

@@

Parker et al. [-@parker2016; -@parker2018; -@parker_information_2018; -@parker2019] have suggested that empirical evidence shows greater school choice at the country level is related to poorer average ability levels, lower aspirations, and paradoxical effects on psychological factors like motivation and self-concept that appears to have negative consequences for all children. They argue that empirical evidence suggests policy should encourage school selection policies that maximize within school heterogeneity. This would require considerable state intervention to achieve and may thus impose unreasonable restrictions on parents' rights to choose. However, it is worth noting that our counterfactual analysis (Table 4) shows that high SES children appear to not suffer to any notable degree, in relation to social skills, by being enrolled in a disadvantaged school.
